Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-09 manuscript/main.tex
computation details to appendix

\begin{definition}[Hybrid inner product between $n$ observations and a single object]\label{def:n_one_inner_product}
Let $\boldsymbol{\xi} = (\psi, \mathbf{s}) \in \mathcal{H}$ be a hybrid object with $\psi \in \mathcal{F}$ and $\mathbf{s} \in \mathbb{R}^p$, and let $\mathbf{a}^{(1)}, \ldots, \mathbf{a}^{(K)}$ denote the basis coefficients of $\psi$ with respect to basis $\Phi^{(1)}, \ldots, \Phi^{(K)}$. Then the inner product between $\boldsymbol{\xi}$ and the $n$ basis-approximated hybrid observations $\widetilde{\mathbf{W}}_1, \ldots, \widetilde{\mathbf{W}}_n$, concatenated as a column vector, is denoted $\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}} \in \mathbb{R}^n$ and computed as
	\begin{equation*}
		\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}
		:=
		\begin{bmatrix}
			\langle \widetilde{\mathbf{W}}_1, \boldsymbol{\xi} \rangle_{\mathcal{H}} \\
			\vdots \\
			\langle \widetilde{\mathbf{W}}_n, \boldsymbol{\xi} \rangle_{\mathcal{H}}
		\end{bmatrix}
		=
		\mathbf{Z} \mathbf{s}
		+
		\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{a}^{(k)}. 
	\end{equation*}
\end{definition}

@@

For the computation of the estimated PLS direction  $\hat{\boldsymbol{\xi}}$, 
we obtain the basis coefficients of its functional part, denoted by $\mathbf{d}^{(1)}, \ldots, \mathbf{d}^{(K)} \in \mathbb{R}^{M}$,
and the scalar part  
$\mathbf{d} \in \mathbb{R}^p$,
by solving linear systems
\begin{equation*}
 \boldsymbol{\Phi}^{(1)}
\mathbf{d}^{(1)} 
=
(\widetilde{\mathbf{C}}^{(1)} \boldsymbol{\Phi}^{(1)})^\top
\mathbf{y},
\ldots
,
\boldsymbol{\Phi}^{(K)}
\mathbf{d}^{(K)} 
=
(\widetilde{\mathbf{C}}^{(K)} \boldsymbol{\Phi}^{(K)})^\top
\mathbf{y},
~\text{and}~
\mathbf{d} := \mathbf{Z}^\top \mathbf{y} .
\end{equation*}
We then normalize these coefficients and scalar parts by the hybrid Hilbert norm (Definition
\ref{def:hilbert_space}),
which can be computed as
\begin{equation*}
\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}} = 
\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}.
\end{equation*}
Therefore, the estimated PLS direction is the tuple $\hat{\boldsymbol{\xi}} \in \mathcal{H}$, defined as
\begin{equation}\label{def:computed_PLS_direction}
	\hat{\boldsymbol{\xi}} :=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
	}
	\biggl(
	\sum_{m=1}^M d^{(1)}_m \phi^{(1)}_m,
	\ldots,
	\sum_{m=1}^M d^{(K)}_m \phi^{(K)}_m,
	\mathbf{d}
	\biggr).
\end{equation}
Finally, the estimated $l$th PLS scores for the $n$ observations,
denoted $\hat{\boldsymbol{\rho}} \in \mathbb{R}^n$
 is computed simultaneously as the hybrid inner product (Definition \ref{def:n_one_inner_product}):
	\begin{equation*}
	\langle \widetilde{\mathbf{W}}, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}}
	=
	\frac{1}{\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}}}
	\biggl(
	\mathbf{Z} \mathbf{d}
	+
	\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{d}^{(k)}
	\biggr)
	. 
\end{equation*}

%	consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
\begin{proposition} \label{Proposition: generalized rayleigh quotient}
	Let $\boldsymbol{\xi} \in \mathcal{H}$ be any unit-norm direction such that its $1$st through $K$th functional components lie in $\operatorname{span}(\Phi^{(1)}), \ldots, \operatorname{span}(\Phi^{(K)})$, respectively.
	Then, the empirical covariance between the projected component $\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the response $Y$, approximated in terms of the basis, is given by
	\begin{equation*}
		\widehat{\textnormal{Cov}} \left( \langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y \right)
		:=
		\frac{1}{n} 
		\mathbf{y}^\top 
		\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}},
	\end{equation*}
where right-hand side notation comes from Definition \ref{def:n_one_inner_product}.
Then the PLS direction 
$	\hat{\boldsymbol{\xi}}$ 
 computed in \eqref{def:computed_PLS_direction} maximizes $\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )$.
	\end{proposition}

@@

	\subsection{Proof of Propositions \ref{Proposition: generalized rayleigh quotient} and \ref{Proposition: Regularized generalized rayleigh quotient} }\label{section:proof:eigenproblem_derivation}

>>>>

@@

For the computation of the estimated PLS direction  $\hat{\boldsymbol{\xi}}$, 
we obtain the basis coefficients of its functional part, denoted by $\mathbf{d}^{(1)}, \ldots, \mathbf{d}^{(K)} \in \mathbb{R}^{M}$,
and the scalar part  
$\mathbf{d} \in \mathbb{R}^p$,
by solving linear systems
\begin{equation*}
 \boldsymbol{\Phi}^{(1)}
\mathbf{d}^{(1)} 
=
(\widetilde{\mathbf{C}}^{(1)} \boldsymbol{\Phi}^{(1)})^\top
\mathbf{y},
\ldots
,
\boldsymbol{\Phi}^{(K)}
\mathbf{d}^{(K)} 
=
(\widetilde{\mathbf{C}}^{(K)} \boldsymbol{\Phi}^{(K)})^\top
\mathbf{y},
~\text{and}~
\mathbf{d} := \mathbf{Z}^\top \mathbf{y} .
\end{equation*}
The estimated PLS direction is the tuple $\hat{\boldsymbol{\xi}} \in \mathcal{H}$, defined with these coefficients and normalized with respect to the 
hybrid Hilbert norm (Definition
\ref{def:hilbert_space})
as follows:
\begin{equation}\label{def:computed_PLS_direction}
	\hat{\boldsymbol{\xi}} :=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
	}
	\biggl(
	\sum_{m=1}^M d^{(1)}_m \phi^{(1)}_m,
	\ldots,
	\sum_{m=1}^M d^{(K)}_m \phi^{(K)}_m,
	\mathbf{d}
	\biggr).
\end{equation}
The details of this hybrid norm computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.
Finally, the estimated $l$th PLS scores for the $n$ observations,
denoted $\hat{\boldsymbol{\rho}} \in \mathbb{R}^n$
 is computed via hybrid inner product. In a matrix multiplication form, it is computed as:
	\begin{align*}
		\hat{\boldsymbol{\rho}}
		&:=
\biggl(
	\langle \widetilde{\mathbf{W}}_1, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}},
	\ldots,
		\langle \widetilde{\mathbf{W}}_n, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}}
\biggr)^\top
\\&	=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
}
	\biggl(
	\mathbf{Z} \mathbf{d}
	+
	\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{d}^{(k)}
	\biggr)
	. 
\end{align*}
The details of this computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.

%	consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
\begin{proposition} \label{Proposition: generalized rayleigh quotient}
	Let $\boldsymbol{\xi} \in \mathcal{H}$ be any unit-norm direction such that its $1$st through $K$th functional components lie in $\operatorname{span}(\Phi^{(1)}), \ldots, \operatorname{span}(\Phi^{(K)})$, respectively.
	Then, the empirical covariance between the projected component $\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the response $Y$, approximated in terms of the basis, is given by
	\begin{equation*}
		\widehat{\textnormal{Cov}} \left( \langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y \right)
		:=
		\frac{1}{n} 
		\sum_{i=1}^n
		y_i
		\langle \widetilde{\mathbf{W}}_i, \boldsymbol{\xi} \rangle_{\mathcal{H}}.
	\end{equation*}
Then the PLS direction 
$	\hat{\boldsymbol{\xi}}$ 
 computed in \eqref{def:computed_PLS_direction} maximizes $\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )$.
	\end{proposition}

@@

	\subsection{Proof of Propositions \ref{Proposition: generalized rayleigh quotient} and \ref{Proposition: Regularized generalized rayleigh quotient} }\label{section:proof:eigenproblem_derivation}
	%
	%
	\begin{align*}
	\widehat{\textnormal{Cov}} (
	\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y 
	) 
	&= 
	\frac{1}{n} \sum \limits_{i=1}^n 
	 y_i
	\langle \widetilde{\mathbf{W}}_i, \boldsymbol{\xi} \rangle_\mathcal{H} \\
	&= 
	 \frac{1}{n} 
	 \sum \limits_{i=1}^n
	 y_i
	   \langle \widetilde{X_i}, \psi \rangle_\mathcal{F}  
	   +  
	   \frac{1}{n} 
	   \sum \limits_{i=1}^n  
	   y_i
	   \langle \mathbf{Z}_i, 
	   \boldsymbol{\theta} 
	   \rangle 
	    \\
	&=  \frac{1}{n} \sum \limits_{i=1}^n  \left\{ \sum \limits_{k=1}^K \int_{\mathcal{T}_k} X_i(t_k) \psi^{(k)}(t_k) dt_k \right\} Y_i +  \frac{1}{n} \sum \limits_{i=1}^n \mathbf{Z}_i^\top \boldsymbol{\theta} Y_i \\
	&=  \frac{1}{n} \sum \limits_{i=1}^n \left[ \sum \limits_{k=1}^K \int_{\mathcal{T}_k} \left\{\sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t) \right\} \left\{\sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t) \right\} dt_k \right] Y_i + \frac{1}{n} \sum \limits_{i=1}^n \mathbf{Z}_i^\top \boldsymbol{\theta} Y_i \\
	&= n^{-1} \widetilde{\mathbf{y}}^\top \left[ \sum \limits_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \left\{ \int_{\mathcal{T}_k} \mathbf{b}^{(k)}(t_k) \mathbf{b}^{(k)}(t_k)^\top dt_k \right\} \mathbf{a}^{(k)} \right] + n^{-1}\widetilde{\mathbf{y}}^\top \mathbf{Z} \boldsymbol{\theta} \\
	&= n^{-1} \widetilde{\mathbf{y}}^\top (\widetilde{C} J \mathbf{a}) + n^{-1}\widetilde{\mathbf{y}}^\top \mathbf{Z} \boldsymbol{\theta} \\
	&= n^{-1} \widetilde{\mathbf{y}}^\top (\widetilde{C}J\mathbf{a} + \mathbf{Z} \boldsymbol{\theta}).
\end{align*}
